---
title: "Online Odyssey Outlet - Data Modelling Script"
author: "Lebintiti Kobe"
format:
    gfm: 
        code-fold: true
        toc: true
        toc-depth: 2
        code-annotations: true
        number-sections: false
        
execute:
  error: false
  warning: false
---

# Load Data 

```{r}
options(scipen = 999)
require(tidyverse)
require(here)
require(skimr)
require(pander)
require(ggthemes)

nov_data <- read_csv(here("data/november.csv.gz"))
dec_data <- read_csv(here("data/december.csv.gz"))

nov_data %>%
    head() %>%
    knitr::kable()
dec_data %>%
    head() %>%
    knitr::kable()
```

Merge november and december data since the have the same data schema then view the data
```{r}
events_data <- bind_rows(nov_data, dec_data)
events_data %>%
    head() %>%
    knitr::kable()
```

## Clean up character data 

-   Convert them to the same casing 
-   Remove whitespaces 

```{r}
events_data <- events_data %>%
    mutate(across(where(is.character), ~ trimws(.) %>% str_to_lower()))
```

## Convert date column into their correct format 

-   Convert date in character format to date format
```{r}
events_data <- events_data %>%
    mutate(across(contains("time"), ~ as_datetime(.))) %>%
    janitor::clean_names()
```

Inspect and skim the data 
```{r}
events_data %>% skim()
```

## Validate the data 

Validate the data to make sure it do not violate business logic : 
 
### Online Stores Business logic : 

User Columns : 

-   

Date Columns : 

 -  Data should be available on all business days dates .
 -  Times of order should always be before delivery times / return times , and so on .

Price Columns : 

-   Prices can be less than zero .
-   Investigate prices that are too high or too low .

Products/Brands/Categories : 

-   Each product ID should be unique to each product . 
-   Each product ID should have one brand/category and so forth . 

Event types : 

-   Usually we should have more product impressions than cart actions , more cart actions than checkouts and so forth . 


Of course , there is so much we can validate for online store data , but for now we focus on the data we have , nothing more nothing less . 

## Products 



## Dates

```{r}
event_data <- events_data %>%
    mutate(event_date = as_date(event_time))
```

### Verify if day has a record 

```{r}
event_data %>%
    group_by(event_date) %>%
    count() %>%
    filter(n < 1)
```


### Check the number of events distribution
```{r}
event_data %>%
    group_by(event_date) %>%
    summarise(n = n()) %>%
    ggplot(aes(x = event_date, y = n, fill = n)) +
    geom_col(show.legend = F) +
    labs(title = "Distribution of events per day", x = "Event Date", y = "") +
    scale_y_continuous(n.breaks = 10, labels = scales::label_number(scale = 0.001, suffix = "K")) +
    theme_excel_new()
```

-   Events are the lowest at the start of november , maybe be the time when the start-up launched . 
-   Events are the highest mid-november , maybe an event was held around this time . 
-   Then after we see a steady rise in events overtime , maybe correspond to normal business growth . 

Enquire about these to figure out what was happening around these times . 
Remember we need to adjust for any artificial boosting of events/sales because these can bias products that were advertised as part of boosting campaigns 


## Price 

### Distribution of prices 
```{r}
event_data %>%
    select(product_id, price) %>%
    distinct() %>%
    ungroup() %>%
    select(price) %>%
    unique() %>%
    ggplot(aes(x = price)) +
    geom_histogram() +
    labs(title = "Distribution of product prices", y = "Number of products", x = "Price") +
    scale_x_continuous(n.breaks = 15, labels = scales::label_currency()) +
    theme_excel_new()
```

-   We have a price range of $0 - $2600 and most products are prices below $600 . 

### Investigate low and high priced products 

```{r}
event_data %>%
    select(-contains("event"), -contains("user")) %>%
    distinct(product_id, .keep_all = T) %>%
    filter(price < 50)

event_data %>%
    select(-contains("event"), -contains("user")) %>%
    distinct() %>%
    filter(price < 2000)
```