---
title: "Online Odyssey Outlet - Data Modelling Script"
author: "Lebintiti Kobe"
format:
    gfm:
      toc: true
      toc-depth: 2
      code-annotations: true
      number-sections: false

execute:
  error: false
  warning: false
editor_options: 
  chunk_output_type: console
---

# Load Libraries

```{r}
options(scipen = 999)
require(tidyverse)
require(here)
require(skimr)
require(ggthemes)
require(kableExtra)
require(arrow)

gc(verbose = F)
```

# Load Data

```{r}
nov_data <- read_csv(here("data/november.csv.gz"))
dec_data <- read_csv(here("data/december.csv.gz"))
```

Merge november and december data since the have the same data schema then view the data

```{r}
events_data <- bind_rows(nov_data, dec_data) %>%
  janitor::clean_names()

rm(nov_data)
rm(dec_data)
gc(verbose = F)

events_data %>% head(n=50) %>% kable()
```

## Clean up character data

-   Convert them to the same casing
-   Remove whitespaces

```{r}
events_data <- events_data %>%
  mutate(across(where(is.character), ~ trimws(.) %>% str_to_lower())) %>%
  mutate(across(contains("id"), ~ as.character(.)))
```

## Convert date column into their correct format

-   Convert date in character format to date format

```{r}
events_data <- events_data %>%
  mutate(across(contains("time"), ~ as_datetime(.)))
```

## Validate the data

Validate the data to make sure it do not violate business logic :

### Online Stores Business logic :

User Columns :

-   Each row of our data should have a session_id

Event types :

-   Usually we should have more product impressions than cart actions , more cart actions than checkouts and so forth .

Date Columns :

-   Data should be available on all business days dates .
-   Times of order should always be before delivery times / return times , and so on .

Price Columns :

-   Prices can be less than zero .
-   Investigate prices that are too high or too low .

Products/Brands/Categories :

-   Each product ID should be unique to each product .
-   Each product ID should have one brand/category and so forth .

Of course , there is so much we can validate for online store data , but for now we focus on concepts that can be found in the data we have , nothing more nothing less .

::: callout-note
In this case if a product has multiple brands , category code and category ID , we assign the most common brand , category code and ID in order to not have multiple values for brand and category columns .

A better way to do things here is to actually assign the most recent brand , category code and ID to products with multiple values in these columns . This makes sense in online store case , so store manager can know which other past values were assign to the most recent , hence use value in those columns .
:::

## User Columns

```{r}
events_data %>%
  select(contains("user")) %>%
  skim()
```

-   We have 27 events records without any user_id . Lets investing these

```{r}
events_data %>%
  filter(is.na(user_session)) %>%
  select(event_type, user_id, price, user_session) %>%
  print(n = 50) %>%
  kable()
```

-   These are all Cart events . These should normally have session_ids , but it is possible , maybe the user lost connection and when they got reconnected they were assigned a new session id

-   For now we keep all these , in order to get true conversion of all out items

## Events types

```{r}
events_data %>%
  janitor::tabyl(event_type) %>%
  kable()
```

-   75% of all our events are cart events , which is to be expected in an online store system . People can add products to cart only to change their mind , or only to buy only those they can afford , mater fact not all items in cart usually get paid for .

## Dates

```{r}
events_data %>%
  select(contains("time")) %>%
  skim()
```

-   Our dates range from 2019-11-01 to 2019-12-31 . Which reassures us that our data was merged correctly .

### Verify if each working day has a record

Unlike physical stores , online store can run even on non-working days .

-   But if we had to find a list of all business days from one date to another , we use the {bizdays} and {timeDate} r package .
-   bizdays::bizseq(from,to,"weekends",holidays) , to find all business days besides weekends but this includes holidays .
-   For holidays , we use timeDate::holidays(from , to , cal = "US") .
-   Then setdiff(bizdays_no_weekends,holidays) , to find all business days not in holidays .

```{r}
events_data <- events_data %>%
  mutate(event_date = as_date(event_time))

events_data %>%
  group_by(event_date) %>%
  count() %>%
  filter(n < 1) %>% head(n=50) %>% kable()
```

-   No results got returned meaning there is no working date without a record / event , therefore we can be a little extra confident on our data

### Check the number of events distribution for each day of our event times

```{r}
events_data %>%
  group_by(event_date) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = event_date, y = n, fill = n)) +
  geom_col(show.legend = F) +
  labs(title = "Distribution of events per day", x = "Event Date", y = "") +
  scale_y_continuous(
    n.breaks = 10,
    labels = scales::label_number(scale = 0.001, suffix = "K")
  ) +
  theme_excel_new()
```

-   Events are the lowest at the start of november , maybe be the time when the start-up launched (It actually the time the start up launched) .
-   Events are the highest mid-november , maybe an event was held around this time .
-   Then after we see a steady rise in events overtime , maybe correspond to normal business growth .

Enquire about these to figure out what was happening around these times . Remember we need to adjust for any artificial boosting of events/sales because these can bias products that were advertised as part of boosting campaigns

## Price

```{r}
events_data %>%
  select(price) %>%
  skim() 
```

-   We have a price range of \\\$0 - \\\$2600 and most products are prices below \$600 . More than 75% of the prices are below \$400

### Investigate low price products

```{r}
events_data %>%
  distinct(product_id, .keep_all = T) %>%
  filter(price <= 0) %>%
  distinct(product_id, .keep_all = T)

zero_price_ids <- events_data %>%
  distinct(product_id, .keep_all = T) %>%
  filter(price <= 0) %>%
  distinct(product_id, .keep_all = T) %>%
  pull(product_id)

events_data %>%
  filter(product_id %in% zero_price_ids) %>%
  arrange(product_id) %>%
  group_by(product_id) %>%
  summarise(
    n_event_type = length(list(unique(event_type))),
    n_brands = length(list(unique(brand)))
  ) %>%
  filter(n_brands > 1 | n_event_type > 1) %>% 
  head(n=50) %>% kable()
```

-   We have around 1400 products with prices of zero , all of them do not have brands codes and they dont have any purchase event . These may products that we dropped from the catalog by some brand which was not selling well .
-   Here we can drop the products since it wont be possible to calculate metrics like Revenue , Conversion , e.t.c

```{r}
events_data <- events_data %>%
  filter(!product_id %in% zero_price_ids)

rm(zero_price_ids)
gc(verbose = F)
```

### Investigate high priced products

```{r}
events_data %>%
  select(-contains("event"), -contains("user")) %>%
  filter(price > 2000) %>%
  janitor::tabyl(category_code) %>%
  arrange(desc(percent)) %>% head(n=50) %>% kable()
```

-   Looking at top categories of high priced items by number of events , the categories listed here are usually high prices , there is nothing too suspicious here .

## Products

Check if each product_id is associated with one brand and category

```{r}
dup_data <- events_data %>%
  group_by(product_id) %>%
  summarise(
    n_brands = n_distinct(brand),
    n_cat_id = n_distinct(category_id),
    n_cat_code = n_distinct(category_code)
  )

dup_ids <- dup_data %>%
  filter(n_brands > 1 | n_cat_id > 1 | n_cat_code > 1) %>%
  pull(product_id)

events_data %>%
  filter(product_id %in% dup_ids) %>%
  arrange(product_id) %>%
  distinct(product_id, category_code, category_id, brand, .keep_all = T) %>% head(n=50) %>% kable()
```

### If a product has two distinct brands / categories but one is NULL , replace NULL with the other brand / category

```{r}
merge_data <- events_data %>%
  filter(product_id %in% dup_ids) %>%
  arrange(product_id) %>%
  distinct(product_id, brand, category_id, category_code) %>%
  group_by(product_id) %>%
  mutate(
    brands = list(brand),
    cat_ids = list(category_id),
    cat_codes = list(category_code)
  ) %>%
  rowwise() %>%
  mutate(
    brands = unique(brands) %>% na.omit() %>% list(),
    cat_ids = unique(brands) %>% na.omit() %>% list(),
    cat_codes = unique(cat_codes) %>% na.omit() %>% list(),
    brand_new = ifelse(
      length(brands) == 1 & length(brands[1]) > 0 & is.na(brand),
      brands[1],
      brand
    ),
    category_id_new = ifelse(
      length(cat_ids) == 1 & length(cat_ids[1]) > 0 & is.na(category_id),
      cat_ids[1],
      category_id
    ),
    category_code_new = ifelse(
      length(cat_codes) == 1 & length(cat_codes[1]) > 0 & is.na(category_code),
      cat_codes[1],
      category_code
    )
  ) %>%
  ungroup()

events_data <- left_join(events_data, merge_data) %>%
  mutate(
    brand = brand_new,
    category_id = category_id_new,
    category_code = category_code_new
  ) %>%
  select(-ends_with("new"), -brands, -cat_ids, -cat_codes)

events_data %>%
  filter(product_id %in% dup_ids) %>%
  arrange(product_id) %>%
  distinct(product_id, category_code, category_id, brand, .keep_all = T) %>% head(n=50) %>% kable()
```

### Check Top Brands and Categories

```{r}
events_data %>%
  janitor::tabyl(brand) %>%
  as_tibble() %>%
  arrange(desc(n)) %>%
  slice_head(n = 20) %>%
  kable(caption = "Top Brands")

events_data %>%
  janitor::tabyl(category_code) %>%
  as_tibble() %>%
  arrange(desc(n)) %>%
  slice_head(n = 20) %>%
  kable(caption = "Top Category Codes")

events_data %>%
  janitor::tabyl(category_id) %>%
  as_tibble() %>%
  arrange(desc(n)) %>%
  slice_head(n = 20) %>%
  kable(caption = "Top Category Ids") 
```

For top brands , find their top categories

```{r}
top_brands <- events_data %>%
  count(brand) %>%
  arrange(desc(n)) %>%
  select(brand)

brand_cat_data <- events_data %>%
  count(brand, category_code) %>%
  arrange(desc(n)) %>%
  left_join(x = top_brands) %>%
  group_by(brand) %>%
  left_join(x = top_brands)

events_data %>%
  count(brand, category_code) %>%
  arrange(desc(n)) %>%
  left_join(x = top_brands) %>%
  group_by(brand) %>%
  slice_max(order_by = n, n = 3) %>%
  left_join(x = top_brands) %>% head(n=50) %>% kable()
```

We see a pattern here , All brands with SMARTPHONE as their second best category all have construction.tools.light as their top categories

We know Brands like Samsung , Apple , Huawei , e.t.c all specialize in smartphone , so construction.tools.light is wrong . We should replace it with smartphones . But we do this for brands that have smartphones as their second best category .

```{r}
events_data %>%
  count(brand, category_code) %>%
  arrange(desc(n)) %>%
  left_join(x = top_brands) %>%
  group_by(brand) %>%
  slice_max(order_by = n, n = 3) %>%
  left_join(x = top_brands) %>%
  group_by(brand) %>%
  mutate(cats = list(category_code)) %>%
  rowwise() %>%
  mutate(
    cats = na.omit(cats) %>% list(),
    category_code_new = ifelse(
      cats[1] == "construction.tools.light" &
        cats[2] == "electronics.smartphone" &
        category_code == "construction.tools.light" &
        length(cats) > 1,
      cats[2],
      category_code
    )
  ) %>%
  select(-n, -cats) %>% head(n=50) %>% kable()
```

Here we see that our fix will work as intended .

```{r}
brand_merge <- brand_cat_data %>%
  group_by(brand) %>%
  mutate(cats = list(category_code)) %>%
  rowwise() %>%
  mutate(
    cats = na.omit(cats) %>% list(),
    category_code_new = ifelse(
      cats[1] == "construction.tools.light" &
        cats[2] == "electronics.smartphone" &
        category_code == "construction.tools.light" &
        length(cats) > 1,
      cats[2],
      category_code
    )
  ) %>%
  select(-n, -cats)

events_data <- events_data %>%
  left_join(y = brand_merge)

rm(merge_data)
rm(dup_data)
rm(dup_ids)
```

```{r}
length(events_data %>% pull(category_code) %>% na.omit())
length(events_data %>% pull(category_code_new) %>% na.omit())
```

Solutions seems to have worked just fine .

```{r}
events_data <- events_data %>%
  mutate(category_code = category_code_new) %>%
  select(-ends_with("new"))
```

```{r}
events_data %>%
  count(brand, category_code) %>%
  arrange(desc(n)) %>%
  left_join(x = top_brands) %>%
  na.omit() %>% head(n=50) %>% kable()

```

The category codes of top brand look fine , there are no suspicious categories that need to be investigated .

```{r}
events_data %>%
  count(brand, category_code, category_id) %>%
  arrange(desc(n)) %>%
  left_join(x = top_brands) %>%
  na.omit() %>% head(n=50) %>% kable()
```

Here , we see that there a unique brand and category combinations with multiple category id .

Now that brands and categories are investigated , lets fix the issue where same product has multiple categories and brands

-   Lets assign top categories and brands to products products with multiple brand and categories , because ideally we should have one category and brand per product .

### Brands

Each product should be associated with one brand , category and category ID

```{r}
dup_data <- events_data %>%
  group_by(product_id) %>%
  summarise(
    n_brands = n_distinct(brand),
    n_cat_id = n_distinct(category_id),
    n_cat_code = n_distinct(category_code)
  )

dup_ids <- dup_data %>%
  filter(n_brands > 1 | n_cat_id > 1 | n_cat_code > 1) %>%
  pull(product_id)

events_data %>%
  filter(product_id %in% dup_ids) %>%
  arrange(product_id) %>%
  distinct(product_id, brand, category_code, category_id, .keep_all = T) %>%
  select(product_id, brand, category_code, category_id) %>% 
  head(n=20) %>% kable()
```

Here we see that product ids have multiple brands , category code and IDS , lets fix this by assigning the most common values to these product IDs

```{r}
merge <- events_data %>%
  count(product_id, brand, category_code, category_id) %>%
  arrange(desc(n)) %>%
  filter(!is.na(product_id)) %>%
  group_by(product_id) %>%
  slice_head(n = 1) %>%
  rename(
    "brand_new" = brand,
    "category_code_new" = category_code,
    "category_id_new" = category_id
  ) %>%
  arrange(desc(n))

events_data <- left_join(events_data, merge) %>%
  mutate(
    brand = brand_new,
    category_code = category_code_new,
    category_id = category_id_new
  ) %>%
  select(-ends_with("new"), -n)

events_data %>%
  filter(product_id %in% dup_ids) %>%
  arrange(product_id) %>%
  distinct(product_id, brand, category_code, category_id, .keep_all = T) %>%
  select(product_id, brand, category_code, category_id) %>%
  head(n=20) %>%
  kable()

events_data %>%
  group_by(product_id) %>%
  summarise(
    n_brands = n_distinct(brand),
    n_cat_id = n_distinct(category_id),
    n_cat_code = n_distinct(category_code)
  ) %>%
  filter(n_brands > 1 | n_cat_id > 1 | n_cat_code > 1) %>%
  kable()
```

Now we see that the product IDs now have one brand , category code and id assigned to them .

### Category Code and ID

In online store databases , category code should be assigned to only one category ID .

```{r}
cat_data <- events_data %>%
  count(category_code, category_id) %>%
  arrange(category_code, desc(n))
cat_data %>% head(n=20) %>% kable()

merge <- cat_data %>%
  group_by(category_code) %>%
  slice_head(n = 1) %>%
  ungroup() %>%
  rename("category_id_new" = category_id)
merge %>% head(n=20) %>% kable()

cat_data <- cat_data %>%
  left_join(y = merge %>% select(-n)) %>%
  select(-n)
cat_data %>% head(n=20) %>% kable()
```

Now each category_code has one ID assigned to it . Now lets merge this with event data in order to fix category codes having multiple category ids issue .

```{r}
events_data %>%
  count(category_code, category_id) %>%
  arrange(category_code, desc(n)) %>% head(n=20) %>% kable()

events_data %>%
  left_join(y = cat_data) %>%
  count(category_code, category_id_new) %>%
  arrange(category_code, desc(n)) %>% head(n=20) %>% kable()
```

Our proposed solution will work as intended

```{r}
events_data <- events_data %>%
  left_join(y = cat_data)

events_data %>% select(category_id) %>% na.omit() %>% nrow()
events_data %>% select(category_id_new) %>% na.omit() %>% nrow()

rm(cat_data)
rm(merge)
rm(dup_data)
rm(dup_ids)
gc(verbose = F)
```

```{r}
events_data <- events_data %>%
  mutate(category_id = category_id_new) %>%
  select(-ends_with("new"))
events_data %>% head(n=20) %>% kable()
```

Verify that each product has one brand , category code and ID .

```{r}
events_data %>%
  group_by(product_id) %>%
  summarise(
    n_brands = n_distinct(brand),
    n_cat_id = n_distinct(category_id),
    n_cat_code = n_distinct(category_code)
  ) %>%
  filter(n_brands > 1 | n_cat_id > 1 | n_cat_code > 1) %>% 
  head(n=20) %>% kable()
```

Now that is our basic data cleaning and data validation done .

## Augment our data

Here we derived extra information from existing columns which we may be useful for us when doing actual analysis . Or we find data to add to current data that we think will be helpful to us when doing our analysis .

Ideas :

-   Finding date , day of week , hour of day and Time without date of our event time column .\
-   Category code values are written in such a way that we can break them down into multiple categories .

These info may be useful for us when doing the actual analyses , in this case the second idea will definitely be useful for us .


### Event time

```{r}
events_data <- events_data %>%
  mutate(
    event_hour = hour(event_time),
    event_day = wday(event_time, label = T, abbr = F),
    event_time_hms = format(event_time, "%H:%M:%S"),
    event_month = month(event_time , label = T , abbr = F)
  )

events_data %>% head(n=20) %>% kable()
```

### Category Code

```{r}
# events_data <- events_data %>%
#   rowwise() %>%
#   mutate(cat_splits = str_split(category_code, "\\.")) %>%
#   ungroup() %>%
#   mutate(length_split = length(cat_splits))
```


The above code is example of bad code . We applying str_split() to all records in our data . Meaning it will be applied to all 7m+ records , we will be computationally intensive . 

The smart of way of doing things here is first extracting all unique category codes , then applying str_split() to the subset then merging the results back into the full dataset . 


```{r}
cat_code_data <- events_data %>% 
  distinct(category_code) %>%
  na.omit() %>%
  rowwise() %>%
  mutate(cat_split = str_split(category_code, "\\."),
  length_split = length(cat_split))
```

```{r}
cat_code_data <- cat_code_data %>%
  ungroup() %>%
  separate(col = category_code , 
  into = paste0("cat_code_",1:max(cat_code_data$length_split)),
  sep = "\\.",
  remove = F) %>%
  select(-contains("split"))

cat_code_data %>% head(n=20) %>% kable()
```

Now we merge the above with our data . 


```{r}
events_data <- events_data %>%
  left_join(y=cat_code_data)

events_data %>% head(n=20) %>% kable()

rm(cat_code_data)
gc(verbose = F)
```

That is our basic data cleaning done . 


## Save the data 

For large datasets like this , it often not advised to save the data as a .csv . 
There are other useful format designed for large data . Example of this is the .parquet format . This format compresses the data , also saves the data types of each column , so when we load the data again in the future we dont start by converting data to correct formats first . 

Popular r package for this format is {arrow} r package . It also allows us to use tidyverse / dplyr syntax when we read in the data .

Use arrow::read_parquet() to read in our .parquet data . 
Use arrow::write_parquet() to save the data in .parquet format .



```{r}
# write_parquet(events_data , 
# "data/events_data_model.gz.parquet",
# compression = "gzip", 
# compression_level = 5)
```



```{r}

```